SI

### 2021-06-10 ms/manuscript.Rmd
Update MS
various small changes after receiving proofs

Comparisons of DT, ETR~max~, and PPFD~95%~ between gametophytes and sporophytes of the same species do not involve phylogeny, so we used a two-sided *t*-test for these.

@@

Desiccation tolerance of gametophytes tended to be lower than sporophytes (`r figure("combined")`), and was significantly lower in `r dt_g_less_than_s` (*t*-test, *P* = `r dt_g_less_than_s_pvals`, respectively).
Only `r dt_g_more_than_s` showed significantly higher DT in gametophytes compared to sporophytes (*t*-test, *P* = `r dt_g_more_than_s_pvals`).

@@

The GLMM including the effect of life stage only was selected as the best-fitting model for ETR~max~, and the combined effect of growth habit plus life stage and their interaction was selected for PPFD~95%~ (`r table("glmm")`).

@@

```{r esm-statement, results = "asis", eval = params$output_type == "ms", purl = FALSE}
# FIXME: add DOI when available
cat("
**Electronic supplementary material**

The online version of this article (DOI) contains supplementary material, which is available to authorized users.
")
```

@@

**`r figure("combined")`** Phylogeny of filmy ferns from Moorea, French Polynesia with comparison of physiological parameters (mean ± 1 SD) between sporophytes (dark grey bars) and gametophytes (light grey bars).
Phylogeny adapted from @Nitta2017; only species with physiological data shown.
Bootstrap support >50% from 100 ML bootstraps indicated at nodes.
Bold letters indicate major clades (H = hymenophylloid clade, T = trichomanoid clade).
Growth habit indicated by colored symbols (blue = low elevation epiphyte, green = high elevation epiphyte, yellow = saxicolous, red = terrestrial).
*Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but are shown as epiphytes to distinguish them from exclusively saxicolous species.
Asterisk indicates significant difference in means (*P* <0.05) between life stages for a particular species (*t*-test); ns indicates no significant difference; lack of ns or asterisk means not enough observations available for that species for *t*-test.
Left bar plot shows recovery (%) of chlorophyll fluorescence (F~v~/F~m~) following 2 day desiccation at –86 MPa (*Abrodictyum dentatum* was desiccated at –38 MPa instead of –86 MPa). 
Middle bar plot shows maximum relative electron transport rate of photosystem II (ETR~max~) (µmol e- m^-2^ s^-1^). 
Right bar plot shows photosynthetic photon flux density at 95% saturation of RETR (PPFD~95%~) (µmol quanta m^-2^ s^-1^).
Absence of a bar indicates measurements not available for gametophytes or sporophytes of that species.

**`r figure("corr")`** Recovery of F~v~/F~m~ (%) following 2 days desiccation at –86 MPa plotted against environmental vapor pressure deficit (VPD; kPa) (**a**, **b**) or range of gametophytes beyond sporophytes (m) (**c**) for gametophytes (**a**, **c**) or sporophytes (**b**). 
Species means ± 1 SD are shown. 
Growth habit indicated by colored symbols (blue = low elevation epiphyte, green = high elevation epiphyte, yellow = saxicolous, red = terrestrial). 
*Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but were treated as epiphytes to distinguish them from exclusively saxicolous species.
*Abrodictyum dentatum* sporophytes were desiccated at –38 MPa instead of –86 MPa.
Range of gametophytes beyond sporophytes calculated as the total elevation that gametophytes exceeded sporophytes at either end of the elevational gradient.
Relationships between recovery and VPD or range of gametophytes beyond sporophytes were analyzed with phylogenetic generalized least squares (PGLS); significant relationships (*P* < 0.05) shown with grey line.

@@

@@

# Part A (left): tree
a <- ggtree(phy) %<+% habit +
  geom_tippoint(aes(color = habit), position = position_nudge(x = 14, y = 0), size = 3.0) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(color = "Growth habit") +
  # There is no way modify the text in geom_treescale (to add units), so make the
  # text disappear by setting fontsize = 0
  geom_treescale(y = 12, x = 2, width = 40, fontsize = 0) +
  # Manually place custom label above scale bar
  annotate("text", x = 22, y = 12, label = "40 my", size = 2.5, vjust = -1) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.04))) +
  theme_void(base_size = 10)

@@

a <- ggplot(plot_data, aes(x = vpd_sporophyte_mean, y = recovery_sporophyte_mean, color = habit)) +
  geom_line(data = vpd_sporophyte_fitted, color = "dark grey") +
  geom_linerange(aes(
    ymin = recovery_sporophyte_mean - recovery_sporophyte_sd,
    ymax = recovery_sporophyte_mean + recovery_sporophyte_sd),
    alpha = 0.3) +
  geom_linerange(aes(
    xmin = vpd_sporophyte_mean - vpd_sporophyte_sd,
    xmax = vpd_sporophyte_mean + vpd_sporophyte_sd),
    alpha = 0.3) +
  geom_point() +
  annotate(
    geom = "richtext", 
    x = Inf, y = -Inf, 
    label = glue("*P* = {vpd_sporophyte_pval}<br>adj. *R*<sup>2</sup> = {vpd_sporophyte_adj_r_squared}"),
    hjust = 1, vjust = 0, size = 3, label.colour = "transparent") +
  scale_y_continuous(
    labels = percent, 
    limits = c(0,1.25)) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(
    y = "Recovery",
    x = "VPD (kPa)",
    color = "Growth habit")

b <- ggplot(plot_data, aes(x = vpd_gametophyte_mean, y = recovery_gametophyte_mean, color = habit)) +
  geom_linerange(aes(
    ymin = recovery_gametophyte_mean - recovery_gametophyte_sd,
    ymax = recovery_gametophyte_mean + recovery_gametophyte_sd),
    alpha = 0.3) +
  geom_linerange(aes(
    xmin = vpd_gametophyte_mean - vpd_gametophyte_sd,
    xmax = vpd_gametophyte_mean + vpd_gametophyte_sd),
    alpha = 0.3) +
  geom_point() +
  scale_y_continuous(
    labels = percent, 
    limits = c(0,1.25)) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(
    y = "Recovery",
    x = "VPD (kPa)",
    color = "Growth habit") +
  guides(color = "none")

c <- ggplot(plot_data, aes(x = gameto_beyond_sporo, y = recovery_gametophyte_mean, color = habit)) +
  geom_linerange(aes(
    ymin = recovery_gametophyte_mean - recovery_gametophyte_sd,
    ymax = recovery_gametophyte_mean + recovery_gametophyte_sd),
    alpha = 0.3) +
  geom_point() +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  scale_y_continuous(labels = percent) +
  labs(
    y = "Recovery",
    x = "Elevation (m)",
    color = "Growth habit") +
  guides(color = "none")

>>>>

Comparisons of DT, ETR~max~, and PPFD~95%~ between gametophytes and sporophytes of the same species do not involve phylogeny, so we used a two-sided *t* test for these.

@@

Desiccation tolerance of gametophytes tended to be lower than sporophytes (`r figure("combined")`), and was significantly lower in `r dt_g_less_than_s` (*t* test, *P* = `r dt_g_less_than_s_pvals`, respectively).
Only `r dt_g_more_than_s` showed significantly higher DT in gametophytes compared to sporophytes (*t* test, *P* = `r dt_g_more_than_s_pvals`).

@@

The GLMM including the effect of life stage only was selected as the best-fitting model for ETR~max~, and the combined effects of growth habit plus life stage and their interaction was selected for PPFD~95%~ (`r table("glmm")`).

@@

```{r esm-statement, results = "asis", eval = params$output_type == "ms", purl = FALSE}
cat("
**Supplementary Information** The online version contains supplementary material available at https://doi.org/10.1007/s10265-021-01318-z.
")
```

@@

**`r figure("combined")`** Phylogeny of filmy ferns from Moorea, French Polynesia with comparison of physiological parameters (mean ± 1 SD) between sporophytes (dark grey bars) and gametophytes (light grey bars).
Phylogeny adapted from @Nitta2017; only species with physiological data shown.
Bootstrap support >50% from 100 ML bootstraps indicated at nodes.
Bold letters indicate major clades (**H** hymenophylloid clade, **T** trichomanoid clade).
Growth habit indicated by colored symbols (blue = low elevation epiphyte, green = high elevation epiphyte, yellow = saxicolous, red = terrestrial).
*Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but are shown as epiphytes to distinguish them from exclusively saxicolous species.
Asterisks indicates significant difference in means (\*\*\* = *P* < 0.001, \*\* = *P* < 0.01, \* = *P* < 0.05) between life stages for a particular species (*t* test); ns indicates no significant difference; lack of ns or asterisk means not enough observations available for that species for *t* test.
Left bar plot shows recovery (%) of chlorophyll fluorescence (F~v~/F~m~) following 2 day desiccation at –86 MPa (*Abrodictyum dentatum* was desiccated at –38 MPa instead of –86 MPa). 
Middle bar plot shows maximum relative electron transport rate of photosystem II (ETR~max~) (µmol e- m^-2^ s^-1^). 
Right bar plot shows photosynthetic photon flux density at 95% saturation of RETR (PPFD~95%~) (µmol quanta m^-2^ s^-1^).
Absence of a bar indicates measurements not available for gametophytes or sporophytes of that species.

**`r figure("corr")`** Recovery of F~v~/F~m~ (%) following 2 days desiccation at –86 MPa plotted against environmental vapor pressure deficit (VPD; kPa) (**a**, **b**) or range of gametophytes beyond sporophytes (m) (**c**) for sporophytes (**a**) or gametophytes (**b**, **c**). 
Species means ± 1 SD are shown. 
Growth habit indicated by colored symbols (blue = low elevation epiphyte, green = high elevation epiphyte, yellow = saxicolous, red = terrestrial). 
*Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but were treated as epiphytes to distinguish them from exclusively saxicolous species.
*Abrodictyum dentatum* sporophytes were desiccated at –38 MPa instead of –86 MPa.
Range of gametophytes beyond sporophytes calculated as the total elevation that gametophytes exceeded sporophytes at either end of the elevational gradient.
Relationships between recovery and VPD or range of gametophytes beyond sporophytes were analyzed with phylogenetic generalized least squares (PGLS); significant relationships (*P* < 0.05) shown with grey line.

@@

# Make tibble of clade labels to plot on tree
clade_labels <-
tibble(
  node = c(
    getMRCA(phy, c("Crepidomanes_bipunctatum", "Callistopteris_apiifolia")),
    getMRCA(phy, c("Hymenophyllum_multifidum", "Hymenophyllum_pallidum"))
  ),
  clade_label = c(
    "T", # trichomanoid clade
    "H" # hymenophylloid clade
  )
)

@@

# Part A (left): tree
a <- ggtree(phy) %<+% habit %<+% clade_labels +
  geom_nodelab(
    aes(label = clade_label), 
    hjust = 1, vjust = 0, geom = "label", 
    label.padding = unit(0.4, "lines"),
    label.size = 0,
    size = 4,
    fill = "transparent",
    fontface = "bold"
    ) +
  geom_tippoint(aes(color = habit), position = position_nudge(x = 14, y = 0), size = 3.0) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(color = "Growth habit") +
  # There is no way modify the text in geom_treescale (to add units), so make the
  # text disappear by setting fontsize = 0
  geom_treescale(y = 12, x = 2, width = 40, fontsize = 0) +
  # Manually place custom label above scale bar
  annotate("text", x = 22, y = 12, label = "40 my", size = 2.5, vjust = -1) +
  scale_x_continuous(expand = expansion(mult = c(0.01, 0.04))) +
  theme_void(base_size = 10)

@@

# a: sporophytes recovery vs. vpd
a <- ggplot(plot_data, aes(x = vpd_sporophyte_mean, y = recovery_sporophyte_mean, color = habit)) +
  geom_line(data = vpd_sporophyte_fitted, color = "dark grey") +
  geom_linerange(aes(
    ymin = recovery_sporophyte_mean - recovery_sporophyte_sd,
    ymax = recovery_sporophyte_mean + recovery_sporophyte_sd),
    alpha = 0.3) +
  geom_linerange(aes(
    xmin = vpd_sporophyte_mean - vpd_sporophyte_sd,
    xmax = vpd_sporophyte_mean + vpd_sporophyte_sd),
    alpha = 0.3) +
  geom_point() +
  annotate(
    geom = "richtext", 
    x = Inf, y = -Inf, 
    label = glue("*P* = {vpd_sporophyte_pval}<br>adj. *R*<sup>2</sup> = {vpd_sporophyte_adj_r_squared}"),
    hjust = 1, vjust = 0, size = 3, label.colour = "transparent") +
  scale_y_continuous(
    labels = percent, 
    limits = c(0,1.25)) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(
    y = "Recovery",
    x = "VPD (kPa)",
    color = "Growth habit")

# b: gametophytes recovery vs. vpd
b <- ggplot(plot_data, aes(x = vpd_gametophyte_mean, y = recovery_gametophyte_mean, color = habit)) +
  geom_linerange(aes(
    ymin = recovery_gametophyte_mean - recovery_gametophyte_sd,
    ymax = recovery_gametophyte_mean + recovery_gametophyte_sd),
    alpha = 0.3) +
  geom_linerange(aes(
    xmin = vpd_gametophyte_mean - vpd_gametophyte_sd,
    xmax = vpd_gametophyte_mean + vpd_gametophyte_sd),
    alpha = 0.3) +
  geom_point() +
  scale_y_continuous(
    labels = percent, 
    limits = c(0,1.25)) +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  labs(
    y = "Recovery",
    x = "VPD (kPa)",
    color = "Growth habit") +
  guides(color = "none")

# c: gametophytes recovery vs G beyond S
c <- ggplot(plot_data, aes(x = gameto_beyond_sporo, y = recovery_gametophyte_mean, color = habit)) +
  geom_linerange(aes(
    ymin = recovery_gametophyte_mean - recovery_gametophyte_sd,
    ymax = recovery_gametophyte_mean + recovery_gametophyte_sd),
    alpha = 0.3) +
  geom_point() +
  scale_color_manual(values = habit_colors, labels = pretty_text) +
  scale_y_continuous(labels = percent) +
  labs(
    y = "Recovery",
    x = "Elevation (m)",
    color = "Growth habit") +
  guides(color = "none")

### 2021-06-10 ms/si.Rmd
Update MS
various small changes after receiving proofs

  ggplot(mapping = aes(x = par)) +
    geom_point(
      data = light_data, 
      aes(y = etr, alpha = outlier)) +
    geom_line(
      data = lc_model_fitted_data, 
      aes(y = etr_fit)) +
    geom_hline(
      data = lc_model_params, 
      aes(yintercept = etr), color = cbPalette[["blue"]]) +
    geom_vline(
      data = lc_model_params, 
      aes(xintercept = par), color = cbPalette[["red"]]) +
    scale_alpha_manual(values = c("TRUE" = 0.2, "FALSE" = 1.0)) +
    labs(
      y = "ETR (µmol m<sup>-2</sup>·s<sup>-1</sup>)", 
      x = "PPFD (µmol m<sup>-2</sup>·s<sup>-1</sup>)", 
      alpha = "Outlier") +
    theme(
      axis.title.x = element_markdown(),
      axis.title.y = element_markdown(),
      strip.text = element_markdown(),
      plot.title = element_markdown(),
      legend.position = "none"
    ) +
    facet_wrap_paginate(
      vars(str_abbrev_sp_pretty(species), light_id), 
      ncol = 4, nrow = 6, page = 1,
      scales = "free", labeller = label_value_2)
}

@@

**`r s_figure("water-all")`**. Relative water content (RWC) of sporophytes after desiccation by duration of desiccation treatment (2 or 15 days).

@@

ggplot(rel_water_species_means_desic, aes(x = rel_water, y = species, group = salt)) +
  geom_errorbar(
    aes(xmin = rel_water - sd, xmax = rel_water + sd),
    color = "grey80",
    width = 0.2,
    position=position_dodge(width=0.5)) +
  geom_point(aes(fill = salt, shape = salt), size = 1.25, position=position_dodge(width=0.5)) +
  scale_y_discrete(labels = str_abbrev_sp_pretty) +
  scale_x_continuous(label = scales::percent) +
  scale_shape_manual(name = "Desiccation intensity",
                     values = salt_shapes,
                     breaks = c("NaCl", "MgNO3",  "LiCl", "Control"),
                     labels = c("-38 MPa", "-86 MPa", "-282 Mpa", "Control")) + 
  scale_fill_manual(name = "Desiccation intensity",
                    values = salt_cols, 
                    breaks = c("NaCl", "MgNO3",  "LiCl", "Control"),
                    labels = c("-38 MPa", "-86 MPa", "-282 Mpa", "Control")) +
  labs(
    x = "Relative water content",
    y = "Species",
    color = "Desiccation\nintensity") +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_markdown(),
    legend.position = "bottom"
  ) +
  facet_grid(vars(dry_time), labeller = as_labeller(c(`2` = "2 days", `15` = "15 days")))
```

**`r s_figure("water-all")`**. Relative water content (RWC) of sporophytes after desiccation by duration of desiccation treatment (2 or 15 days). RWC was not measured for all combinations of species and desiccation treatments; missing points indicate data not collected (not RWC of zero).

@@

**`r s_figure("light-curves")`**. Rapid light response curves of filmy ferns from Moorea, French Polynesia. **a**, sporophytes; **b**, gametophytes. Title of each subplot indicates species followed by light curve ID number assigned to each sample (light curve IDs match those in filmy_light_curves.csv data file). Light curves fit using the equation $y = A(1 - e^{-kx})$, where *y* is relative electron transport rate of photosystem II (ETR), *x* is photosynthetic photon flux density (PPFD), *A* is the asymptote of the curve, and *k* is a slope parameter. Outliers at high PPFD not included in the model indicated in light grey. Blue line indicates maximum modeled relative electron transport rate (ETR~max~). Red line indicates PPFD at 95 % of ETR~max~ (PPFD~95%~).

@@

**`r s_figure("traits-box")`**. Box plots comparing physiological parameters between sporophytes (red) and gametophytes (blue) across growth habits in filmy ferns from Moorea, French Polynesia. Each point represents one species; dark line is median, lower and upper edges correspond to first and third quartiles, and whiskers extend to the most extreme value no further than 1.5 times the inter-quartile range. **a**, ETR~max~ (µmol electrons m^--2^ ·s^--1^); **b** PPFD~95%~ (µmol photons m^--2^ ·s^--1^); **c**, recovery (%) of chlorophyll fluorescence (F~v~/F~m~) following 2 day desiccation at –86 MPa (*Abrodictyum dentatum* was desiccated at –38 MPa instead of –86 MPa). *Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but are treated as epiphytes to distinguish them from exclusively saxicolous species. Growth habit abbreviations: HE, high elevation epiphyte; LE, low elevation epiphyte; S, saxicolous; T, terrestrial.

>>>>

  ggplot(mapping = aes(x = par)) +
    geom_point(
      data = light_data, 
      aes(y = etr, alpha = outlier)) +
    geom_line(
      data = lc_model_fitted_data, 
      aes(y = etr_fit)) +
    geom_hline(
      data = lc_model_params, 
      aes(yintercept = etr), color = cbPalette[["blue"]]) +
    geom_vline(
      data = lc_model_params, 
      aes(xintercept = par), color = cbPalette[["red"]]) +
    scale_alpha_manual(values = c("TRUE" = 0.2, "FALSE" = 1.0)) +
    labs(
      y = "ETR (µmol m<sup>-2</sup> s<sup>-1</sup>)", 
      x = "PPFD (µmol m<sup>-2</sup> s<sup>-1</sup>)", 
      alpha = "Outlier") +
    theme(
      axis.title.x = element_markdown(),
      axis.title.y = element_markdown(),
      strip.text = element_markdown(),
      plot.title = element_markdown(),
      legend.position = "none"
    ) +
    facet_wrap_paginate(
      vars(str_abbrev_sp_pretty(species), light_id), 
      ncol = 4, nrow = 6, page = 1,
      scales = "free", labeller = label_value_2)
}

@@

**`r s_figure("water-all")`**. Relative water content (RWC) of sporophytes in desiccated state by duration of desiccation treatment (2 or 15 days).

@@

ggplot(rel_water_species_means_desic, aes(x = rel_water, y = species, group = salt)) +
  geom_errorbar(
    aes(xmin = rel_water - sd, xmax = rel_water + sd),
    color = "grey80",
    width = 0.2,
    position=position_dodge(width=0.5)) +
  geom_point(aes(fill = salt, shape = salt), size = 1.25, position=position_dodge(width=0.5)) +
  scale_y_discrete(labels = str_abbrev_sp_pretty) +
  scale_x_continuous(label = scales::percent) +
  scale_shape_manual(name = "Desiccation intensity",
                     values = salt_shapes,
                     breaks = c("NaCl", "MgNO3",  "LiCl", "Control"),
                     labels = c("-38 MPa", "-86 MPa", "-282 Mpa", "Control")) + 
  scale_fill_manual(name = "Desiccation intensity",
                    values = salt_cols, 
                    breaks = c("NaCl", "MgNO3",  "LiCl", "Control"),
                    labels = c("-38 MPa", "-86 MPa", "-282 Mpa", "Control")) +
  labs(
    x = "Relative water content",
    y = "Species",
    color = "Desiccation\nintensity") +
  theme_bw() +
  theme(
    panel.grid.minor.y = element_blank(),
    # panel.grid.major.y = element_blank(),
    axis.text.y = element_markdown(),
    legend.position = "bottom"
  ) +
  facet_grid(vars(dry_time), labeller = as_labeller(c(`2` = "2 days", `15` = "15 days")))
```

**`r s_figure("water-all")`**. Relative water content (RWC) of sporophytes in desiccated state (after desiccation but before recovery) by duration of desiccation treatment (2 or 15 days). RWC was not measured for all combinations of species and desiccation treatments; missing points indicate data not collected (not RWC of zero).

@@

**`r s_figure("light-curves")`**. Rapid light response curves of filmy ferns from Moorea, French Polynesia. **a**, sporophytes; **b**, gametophytes. Title of each subplot indicates species followed by light curve ID number assigned to each sample (light curve IDs match those in filmy_light_curves.csv data file available at https://doi.org/10.6084/m9.figshare.14184572). Light curves fit using the equation $y = A(1 - e^{-kx})$, where *y* is relative electron transport rate of photosystem II (ETR), *x* is photosynthetic photon flux density (PPFD), *A* is the asymptote of the curve, and *k* is a slope parameter. Outliers at high PPFD not included in the model indicated in light grey. Blue line indicates maximum modeled relative electron transport rate (ETR~max~). Red line indicates PPFD at 95 % of ETR~max~ (PPFD~95%~).

@@

**`r s_figure("traits-box")`**. Box plots comparing physiological parameters between sporophytes (red) and gametophytes (blue) across growth habits in filmy ferns from Moorea, French Polynesia. Each point represents one species; dark line is median, lower and upper edges correspond to first and third quartiles, and whiskers extend to the most extreme value no further than 1.5 times the inter-quartile range. **a**, ETR~max~ (µmol electrons m^--2^ s^--1^); **b** PPFD~95%~ (µmol photons m^--2^ s^--1^); **c**, recovery (%) of chlorophyll fluorescence (F~v~/F~m~) following 2 day desiccation at –86 MPa (*Abrodictyum dentatum* was desiccated at –38 MPa instead of –86 MPa). *Crepidomanes bipunctatum* and *C. humile* were observed growing as both epiphytes and saxicoles, but are treated as epiphytes to distinguish them from exclusively saxicolous species. Growth habit abbreviations: HE, high elevation epiphyte; LE, low elevation epiphyte; S, saxicolous; T, terrestrial.
